HGP analysis

### 2019-03-05 rick/1KGP.Rmd
update

---
title: "HGP analysis"
author: 
- name: Rick Farouni
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

@@

```{r}
library(Matrix)
library(irlba)
library(threejs)
library(tidyverse)
library(lfa)
library(furrr)
library(broom)
options(future.globals.maxSize= 10000000000)
plan(multicore)
```

# Download data

Download a fast parser program (after compilation you'll have a program called a.out)

```{bash}
wget https://raw.githubusercontent.com/bwlewis/1000_genomes_examples/master/parse.c
cc -O2 parse.c
```

Download 1000 genomes phenotype data file

```{bash}
wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped
```

Download 1000 genomes genotype data file for chr 20

```{bash}
wget ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/release/20130502/ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz

Read just the header of the chromosome file to obtain the sample identifiers

```{r}
ids <- readLines(pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
                      sed -n /^#CHROM/p | 
                      tr '\t' '\n' | 
                      tail -n +10"))
head(ids)
```

Write ids to file

```{r}
writeLines(ids, "ids.txt")
```

@@

```{bash message=FALSE}
for id in `cat ids.txt`
  do
  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.chrom20.*.bam.bas
done
wait
```

# Read bas files

```{r}
read_basfiles <- function(bas_filepath) {

@@

@@

```{r}
write_tsv(bas_dt, "average_quality_chrom21.txt")
```

@@

```{r}
bas_dt <- read_tsv( "average_quality_chrom21.txt")
avg_qual_mean <- bas_dt$avg_qual_mean
```

@@

get_data <- function(){

  p <- pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
          sed /^#/d  |
          cut  -f '10-' |
          ./a.out | 
          cut -f '1-2'")
  x <- read.table(p,
                colClasses=c("integer","integer"), 
                fill=TRUE,
                row.names=NULL)

  data <- sparseMatrix(i=x[,2], j=x[,1], x=1L)

  return(data)

@@

  # compute the first d factors including the intercept
  LF <- lfa(data, d)
  # compute the individual-specific allele frequencies
  data_iaf <- af(data, LF)
  loci <- c(1:dim(data)[1])
  names(loci) <- loci

@@

  data <- get_data()
  data <- t(as.matrix(data[,loci]))
  data_list <- get_af_data(data)

@@

  data_list <- 
    data_list %>%
  slice(1:10) %>%
  transmute(
    locus=locus,
    fit1 = future_map(dataset, fit_model))  %>%
  transmute(
    locus=locus,
    stats = future_map(fit1, c(1)),
    predicted =future_map(fit1, c(2))) %>%
    unnest(stats) %>%
    arrange(-deviance)

@@

```{r}
data_list <- run_workflow (loci=c(1:50000))
data_list
```

```{r}
n <-5
target_loci <- 
  data_list %>% 
  slice(1:n) %>% 
  pull(locus)
target_loci 
```

```{r}
data_list_subset <- 
  data_list %>%
   filter(locus %in% target_loci ) %>%
    mutate(plotted= future_map(predicted, get_plot)) 
```

@@

## Draft

The matrix has 2,504 rows (people) and 1,812,841 columns (variants)

Determine genome-wide goodness-of-fit for a particular value of d by computing SNP-by-SNP goodness-of-fit when compared to population structure.

```{r}
#gof_4 <- model.gof(chr20_mtx, LF, 3)
#LF <- lfa(chr20_mtx, 10)
#gof_10 <- model.gof(chr20_mtx, LF, 3)
#hist(gof_4)
```

data_list
```

```{r}
p_values_dt <-
  data_list  %>% 
  select(tidied) %>% 
  map_depth( 2, "p.value") %>% 
  map_depth( 2, c(2)) %>% # c(1) for the intercept
  unlist(use.names = FALSE) %>% 
  enframe(name = "locus" ,
          value= "p.value")%>%
  mutate(locus= data_list$locus)

p_values_dt
```

```{r}
estimates_dt <-
  data_list  %>% 
  select(tidied) %>% 
  map_depth( 2, "estimate")%>% 
  map_depth( 2, c(2)) %>% # c(1) for the intercept
  unlist(use.names = FALSE) %>% 
  enframe(name = "locus" ,
          value= "beta_hat")%>%
  mutate(locus= data_list$locus)

estimates_dt
```
```{r}
ggplot(p_values_dt)+
  geom_histogram((aes(x=p.value)), bins=100) +
  theme_bw()
```

get a p-value for target specific loci

```{r}
p_values_subset <-
  data_list  %>% 
  filter(locus %in% target_loci) %>% 
  select(tidied) %>% 
  map_depth( 2, "p.value") %>% 
   map_depth( 2, c(2)) %>%
  unlist(use.names = FALSE) %>% 
  enframe(name = "locus" ,
          value= "p.value")%>%
  mutate(locus= target_loci)
p_values_subset 
```

bas_dt <- bind_cols(bas_dt,
                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
```

```{r}
ggplot(bas_dt,
       aes(lf2, lf3, color=Population)) +
  geom_point() +
  theme_bw()
```

### work with a subset

```{r}
chr20_sub <- chr20[,1:50000]
```

# PCA

compute the first three principal component vectors

```{r}
cm <- colMeans(chr20_sub)
p <- irlba(chr20_sub,
           nv=3,
           nu=3,
           tol=0.1,
           center=cm)
```

 add to dataset
```{r}
colnames(p$u) <- paste0("pc",1:3)
bas_dt <- bind_cols(bas_dt, 
                    as_tibble(p$u))
```

```{r}
ggplot(bas_dt,
       aes(pc1, pc2, color=Population)) +
  geom_point() +
  theme_bw() 
```

>>>>

---
title: "HGP analysis"
author: 
- name: Rick Farouni, 
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

@@

```{r}
library(Matrix)
#library(irlba)
#library(threejs)
library(tidyverse)
library(lfa)
library(furrr)
library(broom)
options(future.globals.maxSize= 10000000000)
plan(multicore)
```

# Download metadata

@@

```{bash message=FALSE}
for id in `cat ids.txt`
  do
  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.mapped.*.bam.bas -P ./data/bas
done
wait
```

## Read bas files

```{r read_bas}
read_basfiles <- function(bas_filepath) {

@@

### Preprocess table

@@

```{r}
bas_filepath <- "./data/metadata/average_quality_dt.txt"
write_tsv(bas_dt, bas_filepath)
```

@@

```{r}
bas_dt <- read_tsv(bas_filepath)
avg_qual_mean <- bas_dt$avg_qual_mean
```

@@

@@

  # compute the first d factors including the intercept

  message("running lfa")

  LF <- lfa(data, d)
  # compute the individual-specific allele frequencies

  message("running  af")

  data_iaf <- af(data, LF)
  loci <- c(1:dim(data)[1])
  names(loci) <- loci

  message("to datalist")

@@

  data <- read.bed("./data/plink/ALL_Pops")

  message("plink data loaded")

  data_list <- get_af_data(data)

@@

  message("fitting model")

  data_list <- 
    data_list %>%
  #slice(1:10) %>%
  transmute(
    locus=locus,
    fit1 = future_map(dataset, fit_model))  %>%
  transmute(
    locus=locus,
    stats = future_map(fit1, c(1)),
    predicted =future_map(fit1, c(2))) %>%
    unnest(stats) %>%
    arrange(-deviance)

@@

```{r}
data_list <- run_workflow ()
data_list
```

```{r save_data}
saveRDS(data_list, "./data/preprocessed/data_fit.rds")
```

```{r}
n <-20
target_loci <- 
  data_list %>% 
  slice(1:n) %>% 
  pull(locus)
target_loci 
```

```{r}
data_list_subset <- 
  data_list %>%
   filter(locus %in% target_loci ) %>%
    mutate(plotted= future_map(predicted, get_plot)) 
data_list_subset
```

@@
